Loop Invariant Code Motion in PyPy's Tracing JIT

### 2011-06-16 talk/iwtc11/paper.tex
a motivation example

\subsection{Running Example}
\label{sub:example}

@@

Note that virtuals are only exploded into their attributes when
constructing the arguments of the jump of the preamble. This
explosion can't be repeated when constructing the arguments of the
jump of the peeled loop as it has to mach the first jump. This means that
the objects that was passed as pointers (non virtuals) from the first
iteration to the second (from preamble to peeled loop) also has to be
passed as pointers from the second iteration to the third (from peeled
loop to peeled loop). If one of these objects are virtual 
at the end of the peeled loop they need to be allocated right
before the jump. With the simple objects considered in this paper,
that is not a problem. However in more complicated interpreters such
an allocation might, in combination with other optimizations, lead
to additional variables from the preamble being imported into
the second. This extends both $\hat J$ and $\hat K$, which means that
some care has to be taken, when implementing this, to allow $\hat J$ to
grow while inlining it into $\hat K$. XXX: Maybe we can skip this?

>>>>

\section{Motivation}
\label{sec:Motivation}

To motivate the approach we propose here, let's look at a trivial (unrealistic)
trace which corresponds to an infinite loop:

\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$l_0$($i_{0}$):
$i_1$ = $i_0$ + 1
print($i_1$)
jump($l_0$, $i_0$)
\end{lstlisting}

The first line is a label $l_0$ with argument $i_0$. Every label has a list of
arguments. The \lstinline{print} operation just prints its argument (it is not
an operation that PyPy's tracing JIT really supports, we just use it for this
example). The \lstinline{jump} operation jumps back to the beginning of the
trace, listing the new values of the arguments of the trace. In this case, the
new value of $i_0$ is $i_0$, making it a loop-invariant.

Because $i_0$ is loop-invariant, the addition could be moved out of the loop.
However, we want to get this effect using our existing optimization passes
without changing them too much. To achieve this, we peel one iteration off the
loop before running the optimizations. This peeling gives the following trace:

\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$l_0$($i_{0}$):
$i_1$ = $i_0$ + 1
print($i_1$)
jump($l_1$, $i_0$)

$l_1$($i_{0}$):
$i_2$ = $i_0$ + 1
print($i_2$)
jump($l_1$, $i_0$)
\end{lstlisting}

The iteration of the loop that was peeled off (lines 1-4) is called the
\emph{preamble}, the loop afterwards the \emph{peeled loop}.

Now the optimizer optimizes both of these two iterations of the loop together,
disregarding the \lstinline{jump} and the label in lines 4-6. Doing this, common
subexpression elimination will discover that the two additions are the same, and
replace $i_2$ with $i_1$. This leads to the following trace:

\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$l_0$($i_{0}$):
$i_1$ = $i_0$ + 1
print($i_1$)
jump($l_1$, $i_0$)

$l_1$($i_{0}$):
print($i_1$)
jump($l_1$, $i_0$)
\end{lstlisting}

This trace is malformed, because $i_1$ is used after the label $l_1$ without
being passed there, so we need to add $i_1$ as an argument to the label and pass
it along the \lstinline{jump}s:

\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$l_0$($i_{0}$):
$i_1$ = $i_0$ + 1
print($i_1$)
jump($l_1$, $i_0$, $i_1$)

$l_1$($i_{0}$, $i_1$):
print($i_1$)
jump($l_1$, $i_0$, $i_1$)
\end{lstlisting}

The final result is that the loop-invariant code was moved out of the loop into
the peeled-off iteration. Thus the addition is only executed in the first
iteration, while the result is reused in all further iterations.

This scheme is quite powerful. It allows simple linear optimization passes to
perform loop-aware optimizations, such as loop-invariant code motion without
changing them at all. All that is needed is to peel off one iteration, then
apply simple one-pass optimizations and make sure that the necessary extra
arguments are inserted into the label of the loop itself and the jumps
afterwards. The peeling off of one iteration gives the optimization enough
context to remove operations from the peeled loop, because the optimization
detects that the operation was performed in the preamble already.

% section Motivation (end)

\section{Running Example}
\label{sub:example}

@@

Note that virtuals are only exploded into their attributes when
constructing the arguments of the jump of the preamble. This
explosion can't be repeated when constructing the arguments of the
jump of the peeled loop as it has to match the first jump. This means that
the objects that was passed as pointers (non virtuals) from the first
iteration to the second (from preamble to peeled loop) also has to be
passed as pointers from the second iteration to the third (from peeled
loop to peeled loop). If one of these objects are virtual 
at the end of the peeled loop they need to be allocated right
before the jump. With the simple objects considered in this paper,
that is not a problem. However in more complicated interpreters such
an allocation might, in combination with other optimizations, lead
to additional variables from the preamble being imported into
the second. This extends both $\hat J$ and $\hat K$, which means that
some care has to be taken, when implementing this, to allow $\hat J$ to
grow while inlining it into $\hat K$. XXX: Maybe we can skip this?
